---
title: "Longitudinal Ordinal Data, Conditional Specification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CLMM_conditional}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


Generated data using conditional model - other CLMM Vignettes were using marginal data generation.



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Data Generation

1. Generate longitudinal ordinal data
2. Logistic specification
3. Conditional specification - random intercepts 


## Simulation Study 1, Conditional Data Generation

The conditional data generation does NOT allow you to just fit a model without random intercepts. This is distinct from the marginal models! With marginal models only the SE is affected; with conditional models, you can't ignore the subject effects or else you bias the estimates. So of course the fixed effects only model `clm()` won't handle
missing data correctly, because it can't handle complete data correctly!


### Spoiler: Results, Simulation Study 1
```{r CLMM1_results, eval = F}
                            clm          gee         clmm
0|1                     -1.503910258 -1.504229339 -2.029189347
1|2                      0.012948946  0.012290884  0.009714282
2|3                      1.497568976  1.496833165  2.035309535
GroupGroup_2            -0.001289427  0.005188065 -0.001165598
TimeTime_2              -0.004865262  0.022942622 -0.007264689
TimeTime_3              -0.022885972 -0.011805839 -0.031622008
TimeTime_4               0.011694160  0.003421927  0.014869004
GroupGroup_2:TimeTime_2  0.202491944 -0.202849533  0.270631754
GroupGroup_2:TimeTime_3  0.500235452 -0.501617735  0.675335162
GroupGroup_2:TimeTime_4  0.748884352 -0.749870769  1.012121149
> 
  
 # Marginal means:
> cbind('clm' = colMeans(out.emms.clm), 
+       'clmm' = colMeans(out.emms.clmm))
              clm         clmm
[1,]  0.001289427  0.001165598
[2,] -0.201202517 -0.269466156
[3,] -0.498946025 -0.674169564
[4,] -0.747594924 -1.010955551 

```


### Simulation Study 1 code
```{r CLMM1, eval = F}
rm(list = ls())
gc()

library(SPR)
library(ordinal)
library(multgee)
library(emmeans)

# Initialize output:
out.clm <- vector()
out.clmm <- vector()
out.gee <- vector()
out.emms.clm <- vector()
out.emms.clmm <- vector()
number.repl <- 100


#--------------------------
for (repl in 1:number.repl) {

  set.seed(6252021 + repl)

  sim.out <- SPR::sim_dat_ord_logistic_conditional(N = 300,
                                                   number.groups = 2 ,
                                                   number.timepoints = 4,
                                                   reg.formula = formula( ~ Time + Group + Time*Group),
                                                   Beta = 1,
                                                   thresholds = c(-2, 0, 2),
                                                   subject.var = 2,
                                                   cond.mcar = F,
                                                   Covariate = F)


  dat <- sim.out$dat



# Visualization:
  # Quick barplot, base R plot functions
  # barplot(100*table(dat$Y_comp)/sum(table(dat$Y_comp)),
  #       ylim = c(0, 100), ylab = 'Percentage',
  #       col = 'grey', main = 'Ordinal')



#------------
  # CLM
  mod.clm <- ordinal::clm(as.factor(Y_comp) ~ Group + Time + Group*Time,
                          nAGQ = 5, data= dat)


  # CLMM - Random Effect
  mod.clmm <- ordinal::clmm(as.factor(Y_comp) ~ Group + Time + Group*Time + (1|USUBJID),
                            nAGQ = 5, data= dat)


  # Generalized Estimating Equations:
  dat.gee <- dat
  tmp <- order(dat.gee$id.geepack)
  dat.gee <- dat.gee[tmp, ]
  mod.gee <- multgee::ordLORgee(formula = Y_comp ~ Time + Group + Time*Group,
                                data = dat.gee,
                                id = id.geepack,
                                repeated = Time, LORstr = "uniform")


  # Marginal Means
  # emmeans::emmeans(mod.gee, pairwise ~ Group | Time) NOT AVAILABLE
  emm.clm <- emmeans::emmeans(mod.clm, pairwise ~ Group | Time)
  emm.clmm <- emmeans::emmeans(mod.clmm, pairwise ~ Group | Time)


  # Output
  out.gee <- rbind(out.gee, coef(mod.gee))
  out.clm <- rbind(out.clm, coef(mod.clm))
  out.clmm <- rbind(out.clmm, coef(mod.clmm))
  out.emms.clm  <- rbind(out.emms.clm,
                         as.data.frame(emm.clm$contrast)$estimate)
  out.emms.clmm <- rbind(out.emms.clmm,
                         as.data.frame(emm.clmm$contrast)$estimate)



  cat(paste0('Replication: ', repl, '\n'))

}# end replications

sim.out$Beta
sim.out$thresholds
# All parameters
cbind('clm' = colMeans(out.clm), 
      'gee' = colMeans(out.gee), 
      'clmm' = colMeans(out.clmm))

# Marginal means:
cbind('clm' = colMeans(out.emms.clm), 
      'clmm' = colMeans(out.emms.clmm))

```



## Simulation Study 2, Drop-out

How does the conditional model handle drop-out? Results of simulation study show that it behaves the way you would expect - the CLMM returns unbiased estimates
under MAR drop-out, and biased estimates under MNAR. 

### Spoiler: results of simulation study 2
```{r CLMM2_results, eval = F}
> tab.out[tab.out$Param_type == 'GroupGroup_2:TimeTime_4', ]
       Data              Param_type  Estimate        Bias
41 Complete GroupGroup_2:TimeTime_4 1.0121211  0.01212115
42     MCAR GroupGroup_2:TimeTime_4 1.0204622  0.02046216
43      MAR GroupGroup_2:TimeTime_4 1.0163423  0.01634231
44     MNAR GroupGroup_2:TimeTime_4 0.8633878 -0.13661223
```


### Simulation Study code
```{r CLMM2, eval = F}
rm(list = ls())
gc()

library(SPR)
library(ordinal)
library(emmeans)

# Initialize output:
score.names <- c('Y_comp', 'Y_mcar', 'Y_mar', 'Y_mnar')

out <- data.frame('Param_type' = NA,
                  'Data' = NA,
                  'Estimate' = NA,
                  'Gen_param' = NA)

number.repl <- 100


#--------------------------
  st <- Sys.time()
for (repl in 1:number.repl) {

  set.seed(6252021 + repl)

  sim.out <- SPR::sim_dat_ord_logistic_conditional(
    N = 300,
    number.groups = 2 ,
    number.timepoints = 4,
    reg.formula = formula( ~ Time + Group + Time*Group),
    Beta = 1,
    thresholds = c(-2, 0, 2),
    subject.var = 2,
    cond.mcar = F,
    Covariate = F)


  dat <- sim.out$dat

  dat <- SPR::dropout(dat = dat,
                      type_dropout  = c('mcar', 'mar', 'mnar'),
                      prop.miss = 0.3)


  aggregate(Y_mar ~ Time, FUN = function(x) sum(!is.na(x)),
            na.action = na.pass, data = dat)

  for (score in score.names) {

    # formula:
    mf <- as.formula(paste0('ordered(', score,') ~ Group + Time + Group*Time + (1|USUBJID)'))

    # CLMM - Random Effect
    mod.clmm <- ordinal::clmm(mf, nAGQ = 5, data= dat)

    tmp <- data.frame(
      'Param_type' = names(coef(mod.clmm)),
      'Data' = score,
      'Estimate' = coef(mod.clmm),
      'Gen_param' = c(sim.out$thresholds, as.vector(sim.out$Beta)[-1]))
    # drop the intercept from Beta!

    out <- rbind.data.frame(out, tmp)

    # Marginal Means
    emm.clmm <-
      suppressMessages(
        emmeans::emmeans(mod.clmm, pairwise ~ Group | Time)
      )

    emm <- as.data.frame(emm.clmm$contrasts)

    # Output
    tmp <- data.frame(
      'Param_type' = paste0('emm_', emm$Time),
      'Data' = score,
      'Estimate' = emm$estimate,
      'Gen_param' = as.vector(-1*sim.out$Beta[5:8,]))

    out <- rbind.data.frame(out, tmp)


  }# score loop

  cat(paste0('Replication: ', repl, '\n'))

}# end replications

# Runtime:
et <- Sys.time()
et - st

#---------------------------------------------------------------
# Data Mgmt
out$Data <- ifelse(out$Data == 'Y_comp', 'Complete',
       ifelse(out$Data == 'Y_mcar', 'MCAR',
              ifelse(out$Data == 'Y_mar', 'MAR',
                    ifelse(out$Data == 'Y_mnar', 'MNAR', NA))))

out$Data <- factor(out$Data, levels = c('Complete', 'MCAR', 'MAR', 'MNAR'))
#-------------------------------------------------------------
# Output
tab.out <- aggregate(
  cbind(Estimate,'Bias' = Estimate - Gen_param) ~ Data + Param_type,
  FUN = function(x)  mean(x, na.rm = T),
  na.action = na.pass,
  data = out)

tab.out
tab.out[tab.out$Param_type == 'emm_Time_4', ]
tab.out[tab.out$Param_type == 'GroupGroup_2:TimeTime_4', ]
# g2g


```
