---
title: "Adjusted Relative Risk and Odds Ratios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adjusted_Relative_Risk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

TO DO: Figure out what `epitools::probratio()` is doing such that it lines up exactly
with the adjusted log-binomial model. Is it just running a log-binomial model?

## Overview
Now we extend the OR to RR to include covariates. We can do this with categorical covariates
or continuous covariates. There is an R packge with a function (`epitools::probratio`) that will help you do this, but I want to do it by hand so we all understand what is actually happening. 

Also note that I recently had to compute the adjusted RR for an ordinal regression model. I couldn't use an R library, I had to do it by hand. Once you understand this code below, you'll be able to extend it to ordinal context. 



```{r sim2}
library(SPR)
library(MASS)

N = 1e5 #this should be divisible by however many groups you use!
number.groups <- 2
number.timepoints <- 1
set.seed(9122021)

dat <- data.frame('USUBJID' = rep(paste0('Subject_', formatC(1:N, width = 4, flag = '0')), length.out= N*number.timepoints), 'Group' = rep(paste0('Group_', 1:number.groups), length.out = N*number.timepoints), stringsAsFactors=F)

# Generate a Covariate

# Option 1: Continuous, unbalanced:
# dat$Covariate <- rlnorm(n = N, mean = dat$Group == 'Group_2', sd = 0.25)
# range(dat$Covariate)
# aggregate(Covariate  ~ Group, FUN = function(x) round(mean(x, na.rm = T),2), dat = dat, na.action = na.pass)
# Covariate is unbalanced across groups, could serve as a confound

# Option 2: Categorical, Balanced:
# dat$Z <- sample(x = c(0,1), size = N, replace = T, prob = c(0.5, 0.5))

# Option 3: Categorical, Unbalanced:
dat$Z <- rbinom(n = N, size = 1, prob = 0.4 + 0.4*(dat$Group == 'Group_1'))
table(dat$Z)
xtabs(~ Group + Z, data = dat)


# Create Beta parameters for these design matrix:
X <- model.matrix( ~ Group + Z , data = dat)

# Create Beta
Beta <- matrix(0, nrow = ncol(X), dimnames=list(colnames(X), 'param'))
Beta[] <- c(-0.25, -1, -0.5)

# Matrix multiply:
XB <- X %*% Beta
p <- exp(XB) # prevalence - log link
round(range(p), 2)
#p <- exp(XB)/(1 + exp(XB)) - this is the logistic link

# Dichotomize
dat$Y_binom <- as.vector(1*(p > runif(n = N)))

# Plot
aggregate(Y_binom  ~ Group, FUN = function(x) round(mean(x, na.rm = T),2), dat = dat, na.action = na.pass)
barplot(100*table(dat$Y_binom )/sum(table(dat$Y_binom )), ylim = c(0, 100), ylab = 'Percentage', col = 'grey', main = 'Binomial')


library(ggplot2)
ggplot(data = dat, aes(x= as.factor(Y_binom), fill = Group)) +
      geom_bar(aes(y = (..count..)/sum(..count..)), color="#e9ecef", alpha=0.6, position="dodge", stat="count") +
      scale_y_continuous(labels=scales::percent_format(accuracy = 1)) +
      scale_fill_manual(name = 'Groups', values=c("blue2", "red2")) +
      theme_minimal() +
      theme(legend.position = 'bottom') +
      labs(x = 'Score', y = 'Percentage',
           title = 'Scores with a Binomial Distribution',
           caption = 'Note: Simulated data')

# Observed Data:
xtabs(~ Y_binom + Group, data = dat)
ptab <- prop.table(xtabs(~ Y_binom + Group, data = dat), 2)
```


## Model Fitting - Log Binomial 

This models the relative risk directly. 

```{r mod2}
# Compare to model output:
mod <- glm(Y_binom ~ Group, data = dat, family = binomial(link ='log'))

# Adjusted Relative Risk - log binomial model with covariate (confound)
mod2 <- glm(Y_binom ~ Group + Z,
            start = as.vector(Beta),
            data = dat, family = binomial(link ='log'))

```

## Model Fitting - Logistic Regression 

```{r mod3}
#--------------------------------
# Logistic Regression:
mod.lr <- glm(Y_binom ~ Group + Z, data = dat,
              family = binomial(link ='logit'))

# Now we have the odds ratio from the model, we can compute
# the relative risk using the baseline prevalence.

#
# Can think of several ways of getting the baseline prevalance:
#
# 1. observed rate in control group:
p0_v1 <- ptab['1', 'Group_1']

# 2. Intercept from the logistic regression model:
# Pulled this from Stata function: http://fmwww.bc.edu/repec/bocode/l/logittorisk.ado
odds0 <- exp( coef(mod.lr)['(Intercept)'] )
p0_v2 <- odds0/(1 + odds0)
# but this assumes the covariates are at reference value

# 3. Prediction aligns with intercept:
p0_v3 <- predict(mod.lr, type = 'response',
                 newdata = data.frame('Group' = 'Group_1',
                                      'Z' = 0))

p0_v2 
p0_v3
# So if you use this as your base rate, you are assuming that the other covariate
# is set at the reference level (Z = 0). This may or may not be what you want to do.
# All depends on intended interpretation. 

OR <- exp(coef(mod.lr)['GroupGroup_2'])
P0 <- p0_v2
RR <- OR/( (1-P0) + (P0 * OR) )
RR

# RR computed above assumes Z = 0; to confirm, we compute it this way:
p1 <- predict(mod.lr, type = 'response', newdata = data.frame('Group' = 'Group_2', 'Z' = 0))
p0 <- predict(mod.lr, type = 'response', newdata = data.frame('Group' = 'Group_1','Z' = 0))
p1/p0

# Same value. This is the correct RR if you want to assume the base rate is when Z = 0. 

# Note that it does NOT line up with the generating parameter:
exp(Beta['GroupGroup_2', ])
# Why? because the generating parameter doesn't assume the base rate is when Z = 0

```

## Recover adjusted RR

Can we recover the generating parameter? We want the probability of Y given Group membership, averaged over covariates. This is computing the RR marginalizing over the covariates. So this is the RR specific to the sample characteristics. You've selected a base rate specific to the sample


```{r rr1}
p1i <- predict(mod.lr, type = 'response',
              newdata = data.frame('Group' = 'Group_2',
                                  'Z' = dat$Z))

p0i <- predict(mod.lr, type = 'response',
              newdata = data.frame('Group' = 'Group_1',
                                   'Z' = dat$Z))

mean(p1i)/mean(p0i)

# This aligns with the generating value:
exp(Beta['GroupGroup_2', ])

```

### Note on computation 

https://escholarship.org/content/qt3ng2r0sm/qt3ng2r0sm.pdf

"Standardized measures are constructed by taking averages over C before comparisons (e.g., ratios or differences) across X...Recalling Jensen’s inequality (an average of a nonlinear function does not equal the function applied to the averages), it should not be surprising to find divergences between collapsibility conditions depending on the step at which averaging is done (Samuels, 1981, sec. 3)."


## R package `epitools`

```{r epitools}

library(epitools)
# Function is 'probratio()'
# See documentation - marginalizes over observed covariates
pr <- epitools::probratio(mod.lr, 
                          method='ML', 
                          scale='linear', 
                          start=c(log(mean(dat$Y_binom)), -1,-1))
pr
```

## Summary - All the RR

Let's put it all together and see if we can understand

```{r rr2}

# Generating parameter (true value):
exp(Beta['GroupGroup_2', ]) 

# Observed RR (just using observed proportions):
ptab # Group 1 & Group 2
ptab['1', 'Group_2']/ptab['1', 'Group_1'] 

# RR from unadjusted log-binomial model:
exp(coef(mod)['GroupGroup_2']) 

# RR from Adjusted log-binomial model
exp(coef(mod2)['GroupGroup_2']) 

# RR from adjusted Logistic Regression - Marginalized over covariates 
mean(p1i)/mean(p0i)

# same, but letting `epitools::probratio()` do all the work:
pr


# Logistic Regression (assume base rate is when Z = 0)
RR
p1/p0
# Remember, this isn't wrong *if the research question is, tell me the RR when 
# covariate Z = 0!*


```


## Extension to Ordinal Models

Just going to dump code here - note that the unadjusted lines up very nicely. 

```{r ord1, eval = F}
#-------------
rm(list = ls())
gc()
N = 1e5 #this should be divisible by however many groups you use!
number.groups <- 2
number.timepoints <- 1
set.seed(9132021)

dat <- data.frame(
                  'USUBJID' = rep(paste0('Subject_', formatC(1:N, width = 4, flag = '0')), length.out= N*number.timepoints),
                  'Group' = rep(paste0('Group_', 1:number.groups), length.out = N*number.timepoints),
                  stringsAsFactors=F)


# Create Beta parameters for these design matrix:
X <- model.matrix( ~ Group  , data = dat)


# Create Beta
Beta <- matrix(0, nrow = ncol(X), dimnames=list(colnames(X), 'param'))
Beta[] <- c(0, -0.25)

# Matrix multiply:
XB <- X %*% Beta

# Thresholds:
thr <- c(-2.5, -1.5, -0.5)
eta <- matrix(thr, nrow = nrow(XB), ncol = length(thr), byrow = T) -
  matrix(XB, nrow = nrow(XB), ncol = length(thr), byrow = F)

p <- exp(eta)/(1 + exp(eta))
#p <- exp(eta) # I don't know how to do it this way
range(p)
dat$p <- p
dat$Y <- apply(runif(n = N)  > p, 1, sum)


# Quick barplot, base R plot functions
aggregate(Y  ~ Group, FUN = function(x) round(mean(x, na.rm = T),2), dat = dat, na.action = na.pass)
xtabs( ~ Y + Group, dat = dat, na.action = na.pass)
barplot(100*table(dat$Y)/sum(table(dat$Y)),
        ylim = c(0, 100), ylab = 'Percentage',
        col = 'grey', main = 'Ordinal')

library(ggplot2)
ggplot(data = dat, aes(x= as.factor(Y), fill = Group)) +
      geom_bar(aes(y = (..count..)/sum(..count..)), color="#e9ecef", alpha=0.6, position="dodge", stat="count") +
      scale_y_continuous(labels=scales::percent_format(accuracy = 1)) +
      scale_fill_manual(name = 'Groups', values=c("blue2", "red2")) +
      theme_minimal() +
      theme(legend.position = 'bottom') +
      labs(x = 'Score', y = 'Percentage',
           title = 'Scores with a Binomial Distribution',
           caption = 'Note: Simulated data')

#----
# Key here is that this data is ORDERED

# Observed Data:
xtabs(~ Y + Group, data = dat)
ptab <- prop.table(xtabs(~ Y + Group, data = dat), 2)
# Percentage achieving each category across Groups:
ptab

# Fit ordinal regression
# use the predicted probabilities to compute relative risk
library(ordinal)
mod.clm <- clm(as.factor(Y) ~ Group, data = dat)
summary(mod.clm)
beta <- summary(mod.clm)$beta
alpha <- summary(mod.clm)$alpha


# Group 2:
XB <- matrix(summary(mod.clm)$beta)
thr <- summary(mod.clm)$alpha
eta <- matrix(thr, nrow = nrow(XB), ncol = length(thr), byrow = T) -
  matrix(XB, nrow = nrow(XB), ncol = length(thr), byrow = F)
p1 <- as.numeric(exp(eta)/(1 + exp(eta)))

# Group 1
XB <- matrix(0) # obviously this is more complicated with more predictors!
thr <- summary(mod.clm)$alpha
eta <- matrix(thr, nrow = nrow(XB), ncol = length(thr), byrow = T) -
  matrix(XB, nrow = nrow(XB), ncol = length(thr), byrow = F)
p0 <- (exp(eta)/(1 + exp(eta)))

p0 <- as.numeric(c(0, p0, 1))
p1 <- as.numeric(c(0, p1, 1))
p0 <- diff(p0)
p1 <- diff(p1)
cbind(p0, p1)
ptab
ptab[, 'Group_1']/ptab[, 'Group_2']
p0/p1
#

# > ptab[, 'Group_1']/ptab[, 'Group_2']
#         0         1         2         3
# 0.7987500 0.8373752 0.9105826 1.1056235
# > p0/p1
# [1] 0.7987899 0.8394065 0.9088316 1.1058414
```

### Adjusted Ordinal Model RR

```{r ord2, eval = F}
#-------------
rm(list = ls())
gc()
N = 1e5 #this should be divisible by however many groups you use!
number.groups <- 2
number.timepoints <- 1
set.seed(9152021)

dat <- data.frame(
                  'USUBJID' = rep(paste0('Subject_', formatC(1:N, width = 4, flag = '0')), length.out= N*number.timepoints),
                  'Group' = rep(paste0('Group_', 1:number.groups), length.out = N*number.timepoints),
                  stringsAsFactors=F)

# Unbalanced:
dat$Z <- rbinom(n = N, size = 1, prob = 0.4 + 0.4*(dat$Group == 'Group_1'))
table(dat$Z)
xtabs(~ Group + Z, data = dat)

# Create Beta parameters for these design matrix:
X <- model.matrix( ~ Group + Z , data = dat)


# Create Beta
Beta <- matrix(0, nrow = ncol(X), dimnames=list(colnames(X), 'param'))
Beta[] <- c(0, -0.25, -0.25)

# Matrix multiply:
XB <- X %*% Beta

# Thresholds:
thr <- c(-2, -1, 0)
eta <- matrix(thr, nrow = nrow(XB), ncol = length(thr), byrow = T) -
  matrix(XB, nrow = nrow(XB), ncol = length(thr), byrow = F)

p <- exp(eta)/(1 + exp(eta))
#p <- exp(eta)
dat$p <- p
dat$Y <- apply(runif(n = N)  > p, 1, sum)


# Quick barplot, base R plot functions
aggregate(Y  ~ Group, FUN = function(x) round(mean(x, na.rm = T),2), dat = dat, na.action = na.pass)
xtabs( ~ Y + Group, dat = dat, na.action = na.pass)
barplot(100*table(dat$Y)/sum(table(dat$Y)),
        ylim = c(0, 100), ylab = 'Percentage',
        col = 'grey', main = 'Ordinal')

library(ggplot2)
ggplot(data = dat, aes(x= as.factor(Y), fill = Group)) +
      geom_bar(aes(y = (..count..)/sum(..count..)), color="#e9ecef", alpha=0.6, position="dodge", stat="count") +
      scale_y_continuous(labels=scales::percent_format(accuracy = 1)) +
      scale_fill_manual(name = 'Groups', values=c("blue2", "red2")) +
      theme_minimal() +
      theme(legend.position = 'bottom') +
      labs(x = 'Score', y = 'Percentage',
           title = 'Scores with a Binomial Distribution',
           caption = 'Note: Simulated data')

#----
# Key here is that this data is ORDERED

# Observed Data:
xtabs(~ Y + Group, data = dat)
ptab <- prop.table(xtabs(~ Y + Group, data = dat), 2)
# Percentage achieving each category across Groups:
ptab

# Fit ordinal regression
# use the predicted probabilities to compute relative risk
library(ordinal)
mod.clm <- clm(as.factor(Y) ~ Group + Z, data = dat)
summary(mod.clm)
beta <- summary(mod.clm)$beta
alpha <- summary(mod.clm)$alpha


# Group 2:
X1 <- cbind('Group' = rep(1, N), 'Z' = dat$Z)
BB <- matrix(summary(mod.clm)$beta)
XB <- X1 %*% BB
#XB <- matrix(summary(mod.clm)$beta)
thr <- summary(mod.clm)$alpha
eta <- matrix(thr, nrow = nrow(XB), ncol = length(thr), byrow = T) -
  matrix(XB, nrow = nrow(XB), ncol = length(thr), byrow = F)
p1 <- exp(eta)/(1 + exp(eta))

# Group 1
X0 <- cbind('Group' = rep(0, N), 'Z' = dat$Z)
BB <- matrix(summary(mod.clm)$beta)
XB <- X0 %*% BB
#XB <- matrix(0) # obviously this is more complicated with more predictors!
thr <- summary(mod.clm)$alpha
eta <- matrix(thr, nrow = nrow(XB), ncol = length(thr), byrow = T) -
  matrix(XB, nrow = nrow(XB), ncol = length(thr), byrow = F)
p0 <- exp(eta)/(1 + exp(eta))

p0 <- cbind(0, p0, 1)
p1 <- cbind(0, p1, 1)
p0 <- t(apply(p0, 1, diff))
p1 <- t(apply(p1, 1, diff))
p0 <- colMeans(p0)
p1 <- colMeans(p1)
ptab
cbind(p0, p1)
ptab[, 'Group_1']/ptab[, 'Group_2']
p0/p1

#------------------------------------
# > ptab[, 'Group_1']/ptab[, 'Group_2']
#         0         1         2
# 0.8732260 0.8552308 1.1510231

# > p0/p1
# [1] 0.6966564 0.8638131 1.2358708

```


