---
title: "How to Handle Missing Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing_Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Example 1: MCAR
MCAR (standard interpretation of MCAR - descriptive statistics sufficient!)


```{r mcar_1, eval = F}

library(MASS)
library(emmeans)
library(nlme)
library(SPR) 
  
  out <- sim_dat(N = 2000, 
                 number.groups = 2 , 
                 number.timepoints = 4, 
                 reg.formula = formula( ~ Time + Group + Time*Group),
                 Beta = 2,
                 corr = 'ar1', 
                 cor.value = 0.8, 
                 var.values = 2)
                
  
  dat <- out$dat
  dat <- dropout(dat = dat, 
                 type_dropout  = c('mcar'), 
                 prop.miss = 0.3)
  
 
#  OLS Regression Model 
  mod.ols1 <- lm(Y_comp ~ Time + Group + Time*Group, 
                 data = dat)

  mod.ols2 <- lm(Y_mcar ~ Group + Time + Group*Time, 
                            data = dat)

#   MMRM 
  library(nlme)
  mod.gls1 <- gls(Y_comp ~ Group + Time + Group*Time,
                  data = dat,
                  correlation = corSymm(form = ~ 1 | USUBJID),    #  unstructured correlation
                  weights = varIdent(form = ~ 1 | Time),          #  freely estimate variance at subsequent timepoints
                  na.action = na.exclude)


  mod.gls2 <- gls(Y_mcar ~ Group + Time + Group*Time,
                  data = dat,
                  correlation = corSymm(form = ~ 1 | USUBJID),    #  unstructured correlation
                  weights = varIdent(form = ~ 1 | Time),          #  freely estimate variance at subsequent timepoints
                  na.action = na.exclude)
  
  
  # Compute marginal mean contrasts
  library(emmeans)
  mod.emms.ols1 <- emmeans(mod.ols1, pairwise ~ Group | Time, adjust = 'none',  mode = 'df.error')
  mod.emms.ols2 <- emmeans(mod.ols2, pairwise ~ Group | Time, adjust = 'none',  mode = 'df.error')

 # mod.emms.gls1 <- emmeans(mod.gls1, pairwise ~ Group | Time, adjust = 'none',  mode = 'satterthwaite')
  mod.emms.gls1 <- emmeans(mod.gls1, pairwise ~ Group | Time, adjust = 'none',  mode = 'df.error')
 # mod.emms.gls2 <- emmeans(mod.gls2, pairwise ~ Group | Time, adjust = 'none',  mode = 'satterthwaite')
  mod.emms.gls2 <- emmeans(mod.gls2, pairwise ~ Group | Time, adjust = 'none',  mode = 'df.error')
# Analytical Satterthwaite method not available; using appx-satterthwaite
# Error: Can't estimate Satterthwaite parameters.
#   Try adding the argument 'mode = "df.error"'
 
```

## MCAR: Examine output

```{r mcar_2, eval = F}

  out$Beta
  # Look at drop-out:
  aggregate(Y_mcar ~ Time, FUN = function(x) mean(is.na(x)), dat = dat, na.action = na.pass)
  aggregate(Y_mcar ~ Group + Time, FUN = function(x) mean(is.na(x)), dat = dat, na.action = na.pass)
  # Approximately equal across both treatment arms - already know there's no way for the drop-out to bias
  # the treatment arm comparisons!
  
  # Descriptive Statistics 
  # Sufficient with MCAR drop out - both complete and missing align, yield unbiased estimates:
  aggregate(Y_comp ~ Group + Time, FUN = function(x) round(mean(x, na.rm = T),5), dat = dat, na.action = na.pass)
  aggregate(Y_mcar ~ Group + Time, FUN = function(x) round(mean(x, na.rm = T),5), dat = dat, na.action = na.pass)

  # OLS likewise recovers unbiased parameters using complete data AND with MCAR drop-out
  mod.emms.ols1$contrasts
  mod.emms.ols2$contrasts
  # Here we are only running a single replication to illustrate the unbiased estimates
  # if we ran a full simulation study, we would see that RMSE increases under MCAR drop-out,
  # resulting from the reduction in sample size - no getting away from that!
  mod.emms.gls1$contrasts
  mod.emms.gls2$contrasts
  # Can see that the correctly specified model - MMRM - yields the same estimates
#
```

## Test MCAR Assumption
Does your score at a previous timepoint predict your score at the next timepoint? Let's make our life easy and just do the following:

```{r mcar_3, eval = F}


dat1 <- dat[dat$Time == 'Time_3', 'Y_mcar']
dat2 <- dat[dat$Time == 'Time_4', 'Y_mcar']
dropped.out <- !is.na(dat1) & is.na(dat2) # these subjects dropped out at timepoint 4
table(dropped.out)

# Does the score at timepoint 3 predict drop-out at timepoint 4?
mod <- glm(dropped.out ~ dat1, family = 'binomial')
summary(mod)
# This is evidence suggesting that the dropout may be MCAR
# Later we will re-compute this test on MAR dropout


```


# Example 2: MAR Drop-out
MAR with correlation = 0.8  (standard interpretation of MAR dropout - descriptive statistics, OLS not sufficient!)

```{r mar_1, eval = F}

  dat <- out$dat
  dat <- dropout(dat = dat, 
                 type_dropout  = c('mar'), 
                 prop.miss = 0.3)
 
# Test for MCAR
# Does your score at a previous timepoint predict your score at the next timepoint?
dat1 <- dat[dat$Time == 'Time_3', 'Y_mar']
dat2 <- dat[dat$Time == 'Time_4', 'Y_mar']
dropped.out <- !is.na(dat1) & is.na(dat2) # these subjects dropped out at timepoint 4
table(dropped.out)

# Does the score at timepoint 3 predict drop-out at timepoint 4?
mod <- glm(dropped.out ~ dat1, family = 'binomial')
summary(mod)  
# Reject null hypothesis; timepoint 3 predicts dropout at next timepoint
# Evidence against MCAR drop-out



#  OLS Regression Model 
  mod.ols2 <- lm(Y_mar ~ Group + Time + Group*Time, data = dat)

# MMRM
  mod.gls2 <- gls(Y_mar ~ Group + Time + Group*Time,
                  data = dat,
                  correlation = corSymm(form = ~ 1 | USUBJID),    #  unstructured correlation
                  weights = varIdent(form = ~ 1 | Time),          #  freely estimate variance at subsequent timepoints
                  na.action = na.exclude)

# Marginal Contrasts:
  mod.emms.ols2 <- emmeans(mod.ols2, pairwise ~ Group | Time, adjust = 'none',  mode = 'df.error')
  mod.emms.gls2 <- emmeans(mod.gls2, pairwise ~ Group | Time, adjust = 'none',  mode = 'df.error')
  
  # Examine output:
  out$Beta
  # Look at drop-out:
  aggregate(Y_mar ~ Time, FUN = function(x) mean(is.na(x)), dat = dat, na.action = na.pass)
  aggregate(Y_mar ~ Group + Time, FUN = function(x) mean(is.na(x)), dat = dat, na.action = na.pass)
  # Substantial differential rates of drop-out across the two treatment arms

  # Descriptive Statistics 
  # Substantial bias in the descriptive statistics when MAR dropout
  aggregate(Y_comp ~ Group + Time, FUN = function(x) round(mean(x, na.rm = T),2), dat = dat, na.action = na.pass)
  aggregate(Y_mar ~ Group + Time, FUN = function(x) round(mean(x, na.rm = T),2), dat = dat, na.action = na.pass)
  # Do the arithmetic with the decsriptive statistics, compare to OLS:
  
  # OLS likewise has substantial bias with MAR dropout:
  mod.emms.ols1$contrasts
  mod.emms.ols2$contrasts
  # Notice that the OLS contrasts are equal to those using descriptive statistics

  # MMRM is correctly specified model:
  mod.emms.gls1$contrasts
  mod.emms.gls2$contrasts
  # MMRM unbiased with MAR drop-out
  # In this single replication, we see very little bias
  # Not bad for 30% drop-out!
```

# Example 3: MNAR Drop-out
MNAR with correlation = 0.8 (standard interpretation of MNAR - nothing works!)

```{r mnar_1, eval = F}

 dat <- out$dat
  dat <- dropout(dat = dat, 
                 type_dropout  = c('mnar'), 
                 prop.miss = 0.3)

# Test for MCAR 
# Does your score at a previous timepoint predict your score at the next timepoint?
dat1 <- dat[dat$Time == 'Time_3', 'Y_mnar']
dat2 <- dat[dat$Time == 'Time_4', 'Y_mnar']
dropped.out <- !is.na(dat1) & is.na(dat2) # these subjects dropped out at timepoint 4
table(dropped.out)

# Does the score at timepoint 3 predict drop-out at timepoint 4?
mod <- glm(dropped.out ~ dat1, family = 'binomial')
summary(mod)  
# Reject null hypothesis; timepoint 3 predicts dropout at next timepoint
# Evidence against MCAR drop-out

 
#  OLS Regression Model 
  mod.ols2 <- lm(Y_mnar ~ Group + Time + Group*Time, 
                            data = dat)

# MMRM
  mod.gls2 <- gls(Y_mnar ~ Group + Time + Group*Time,
                  data = dat,
                  correlation = corSymm(form = ~ 1 | USUBJID),    #  unstructured correlation
                  weights = varIdent(form = ~ 1 | Time),          #  freely estimate variance at subsequent timepoints
                  na.action = na.exclude)  
  
# Marginal Contrasts:
  mod.emms.ols2 <- emmeans(mod.ols2, pairwise ~ Group | Time, adjust = 'none',  mode = 'df.error')
  mod.emms.gls2 <- emmeans(mod.gls2, pairwise ~ Group | Time, adjust = 'none',  mode = 'df.error')
 
  # Examine output:
  out$Beta
  # Look at drop-out:
  aggregate(Y_mnar ~ Time, FUN = function(x) mean(is.na(x)), dat = dat, na.action = na.pass)
  aggregate(Y_mnar ~ Group + Time, FUN = function(x) mean(is.na(x)), dat = dat, na.action = na.pass)
  # Substantial differential rates of drop-out across the two treatment arms

  # Descriptive Statistics 
  # Substantial bias in the descriptive statistics when MAR dropout
  aggregate(Y_comp ~ Group + Time, FUN = function(x) round(mean(x, na.rm = T),2), dat = dat, na.action = na.pass)
  aggregate(Y_mnar ~ Group + Time, FUN = function(x) round(mean(x, na.rm = T),2), dat = dat, na.action = na.pass)

  # OLS likewise has substantial bias with MAR dropout:
  mod.emms.ols1$contrasts
  mod.emms.ols2$contrasts
  # 25% bias!

  # MMRM is correctly specified model:
  mod.emms.gls1$contrasts
  mod.emms.gls2$contrasts
  # MMRM still biased with MNAR drop-out
  # However, it's not AS BAS as OLS and descriptive statistics
``` 


